---
title: Quality assessment and processing of human DLPFC data
author: 
- name: Pere Moles
  affiliation:
  - MSc Data Science student, Universitat Oberta de Catalunya
  email: pere.moles@uoc.edu
date: "`r format(Sys.time(), '%B %e, %Y')`"
abstract: >
  Here we perform a first exploratory analysis and quality assessment of the 10x Visium human dorsolateral pre-frontal cortex (DLPFC) data, and process it for its downstream analysis.
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    fig_captions: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: _bibliography.bib
---

```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(kableExtra)


knitr::opts_chunk$set(
  collapse=TRUE,
  comment="",
  fig.align="center",
  fig.wide=TRUE,
  cache=TRUE,
  cache.path="_cache/__QA_hdlpfc",
  cache.extra=R.version.string,
  autodep=TRUE
)
```

# Download data

We start by downloading the data, following the instructions at
http://research.libd.org/spatialLIBD/articles/spatialLIBD.html

```{r, message=FALSE}
library(ExperimentHub)
library(SpatialExperiment)
library(spatialLIBD)

ehub <- ExperimentHub()
spe <- fetch_data(type="spe", eh=ehub)
dim(spe)
spe
```
This is the row data.

```{r}
rowData(spe)
```
This is the colum data. We have multiple samples and a few thousand spots per
sample.

```{r}
colData(spe)
table(colData(spe)$sample_id)
```
For simplicity, we are going to analyze here only the sample with
identifier `sample_151673` to match the analysis shown at
http://research.libd.org/spatialLIBD/articles/spatialLIBD.html

```{r}
spe <- spe[, colData(spe)$sample_id %in% "151673"]
dim(spe)
```

There is an indicator column called `in_tissue` for identifying spots
overlapping the tissue, but apparently all spots have such a condition.

```{r}
stopifnot(all(colData(spe)$in_tissue))
```

The column data includes a column called `spatialLIBD` with anatomically
annotated groups of spots (six cortical layers and white matter). This
annotation is apparently missing for a small number of spots, we are
going to discard those spots as well.

```{r}
table(colData(spe)$spatialLIBD, useNA="always")
spe <- spe[, !is.na(colData(spe)$spatialLIBD)]
dim(spe)
table(colData(spe)$spatialLIBD, useNA="always")
```


This is the internal column metadata.

```{r}
int_colData(spe)
```

# Quality assessment

Identify mitochondrial (MT) genes, here the column with the gene HUGO symbols
is called `gene_name`.

```{r}
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
table(is_mito)
rowData(spe)$gene_name[is_mito]
```
Add quality control metrics.

```{r, message=FALSE}
library(scuttle)
library(ggspavis)
library(patchwork)

spe <- addPerCellQC(spe, subsets=list(mito=is_mito))
```
Explore sequencing depth in UMI counts per spot. Figure
\@ref(fig:hdlpfctissueUMIcntxspot) shows in the tissue image the spatial
pattern of sequencing depth in UMI counts per spot.

```{r hdlpfctissueUMIcntxspot, fig.height=5, fig.width=5, out.width="600px", fig.cap="Spatial pattern of sequencing depth per spot in UMI counts.", dpi=100}
## image seems to be reversed and for this reason we set x_coord="pxl_col_in_fullres"
## and y_coord="pxl_row_in_fullres"
plotVisium(spe, x_coord="pxl_col_in_fullres",
           y_coord="pxl_row_in_fullres",
           annotate="sum", point_size=1,
           pal=c("skyblue", "darkred"), facets=NULL)
```

Figure \@ref(fig:hdlpfcUMIcntxspot)a
shows the frequency distribution of UMI counts per spot and panels (b) to (d)
show the spatial distribution of discarded spots at different cutoffs of
sequencing depth (2000, 1000 and 700).

```{r hdlpfcUMIcntxspot, fig.height=7, fig.width=7, out.width="800px", fig.cap="Sequencing depth per spot in UMI counts. (a) Frequency distribution of UMI counts, where vertical lines indicate possible cutoff UMI count values below which spots would be discarded. (b) to (d) spatial distribution of spots discarded using the ppreviously considered sequencing depth cutoff values."}
hp <- ggplot(data=colData(spe), aes(x=sum)) +
        geom_histogram(bins=40) +
        labs(x="UMI counts per spot", y="Frequency", tag="a") +
        geom_vline(aes(xintercept=2000), col="darkgreen") +
        geom_vline(aes(xintercept=1000), col="darkorange") +
        geom_vline(aes(xintercept=700), col="darkred")
colData(spe)$qc_lib_size <- colData(spe)$sum < 2000
qp2k <- plotSpotQC(spe, plot_type="spot", annotate="qc_lib_size") +
          theme(legend.position="none") + labs(tag="b")
colData(spe)$qc_lib_size <- colData(spe)$sum < 1000
qp1k <- plotSpotQC(spe, plot_type="spot", annotate="qc_lib_size") +
          theme(legend.position="none") + labs(tag="c")
colData(spe)$qc_lib_size <- colData(spe)$sum < 700
qp700 <- plotSpotQC(spe, plot_type="spot", annotate="qc_lib_size") +
          theme(legend.position="none") + labs(tag="d")
(hp | qp2k) / (qp1k | qp700)
```

Explore number of genes with non-zero UMI counts per spot. Figure
\@ref(fig:hdlpfcnonzeroUMIcntxspot)a shows the frequency distribution of genes
per spot with non-zero UMI counts, while panels (b) do (d)
show the spatial distribution of discarded spots at different cutoffs of
sequencing depth (1000, 750 and 500).


```{r hdlpfcnonzeroUMIcntxspot, fig.height=7, fig.width=7, out.width="800px", fig.cap="Non-zero UMI counts per spot. (a) Frequency distribution of non-zero UMI counts per spot, where vertical lines indicate possible cutoff values in the number of genes with non-zero UMI counts, below which spots would be discarded. (b) to (d) spatial distribution of spots discarded using the ppreviously considered sequencing depth cutoff values."}
hp <- ggplot(data=colData(spe), aes(x=detected)) +
        geom_histogram(bins=40) +
        labs(x="Genes with non-zero UMI counts per spot",
             y="Frequency", tag="a") +
        geom_vline(aes(xintercept=1000), col="darkgreen") +
        geom_vline(aes(xintercept=750), col="darkorange") +
        geom_vline(aes(xintercept=500), col="darkred")
colData(spe)$qc_detected <- colData(spe)$detected < 1000
qp1k <- plotSpotQC(spe, plot_type="spot", annotate="qc_detected") +
          theme(legend.position="none") + labs(tag="b")
colData(spe)$qc_detected <- colData(spe)$detected < 750
qp750 <- plotSpotQC(spe, plot_type="spot", annotate="qc_detected") +
          theme(legend.position="none") + labs(tag="c")
colData(spe)$qc_detected <- colData(spe)$detected < 500
qp500 <- plotSpotQC(spe, plot_type="spot", annotate="qc_detected") +
          theme(legend.position="none") + labs(tag="d")
(hp | qp1k) / (qp750 | qp500)
```
Figure \@ref(fig:hdlpfcMTUMIcntxspot) shows the proportion of MT UMI counts per
spot.

```{r hdlpfcMTUMIcntxspot, fig.height=7, fig.width=7, out.width="800px", fig.cap="Mitochondrial UMI counts per spot. (a) Frequency distribution of the proportion of mitochondrial UMI counts per spot, where vertical lines indicate possible cutoff values in the proportion of UMI counts from mitochondrial genes, above which spots would be discarded. (b) to (d) spatial distribution of spots discarded using the ppreviously considered sequencing depth cutoff values."}
hp <- ggplot(data=colData(spe), aes(x=subsets_mito_percent)) +
        geom_histogram(bins=40) +
        labs(x="Proportion of mitochondrial UMI counts per spot",
             y="Frequency", tag="a") +
        geom_vline(aes(xintercept=25), col="darkgreen") +
        geom_vline(aes(xintercept=27), col="darkorange") +
        geom_vline(aes(xintercept=29), col="darkred")
colData(spe)$qc_mito <- colData(spe)$subsets_mito_percent > 25
qp25 <- plotSpotQC(spe, plot_type="spot", annotate="qc_mito") +
          theme(legend.position="none") + labs(tag="b")
colData(spe)$qc_mito <- colData(spe)$subsets_mito_percent > 27
qp27 <- plotSpotQC(spe, plot_type="spot", annotate="qc_mito") +
          theme(legend.position="none") + labs(tag="c")
colData(spe)$qc_mito <- colData(spe)$subsets_mito_percent > 29
qp29 <- plotSpotQC(spe, plot_type="spot", annotate="qc_mito") +
          theme(legend.position="none") + labs(tag="d")
(hp | qp25) / (qp27 | qp29)
```
We use the more lenient cutoff values in the previous three
QC metrics, which lead to the following amount of discarded
spots per metric.

```{r}
colData(spe)$discard <- colData(spe)$qc_lib_size |
                        colData(spe)$qc_detected |
                        colData(spe)$qc_mito
table(colData(spe)$discard)
```
Figure \@ref(fig:hdlpfcdiscardedspots) below shows the spatial pattern of the
discarded spots.

```{r hdlpfcdiscardedspots, fig.height=4, fig.width=8, out.width="800px", fig.cap="Spatial pattern of discarded spots. (a) Discarded spots highlighted in red in the spot grid. (b) Discarded spots highlighted in blue in the tissue plot."}
pqc <- plotSpotQC(spe, plot_type="spot", annotate="discard") +
         theme(legend.position="none") + labs(tag="a")
## image seems to be reversed and for this reason we set x_coord="pxl_col_in_fullres"
## and y_coord="pxl_row_in_fullres"
pts <- plotVisium(spe, x_coord="pxl_col_in_fullres", facets=NULL,
                  y_coord="pxl_row_in_fullres", annotate="discard") +
         labs(tag="b") + theme(legend.position="none")
(pqc | pts)
```
Finally, we filter out these low-quality spots.

```{r}
dim(spe)
spe <- spe[, !colData(spe)$discard]
dim(spe)
```

# Filter lowly-expressed genes

Figure \@ref(fig:hdlpfcUMIcntxgene) shows the empirical distribution of UMI
counts per gene.

```{r hdlpfcUMIcntxgene, fig.height=4, fig.width=6, out.width="500px", fig.cap="UMI counts per gene. The vertical red line shows a cutoff value below which we discard lowly-expressed genes."}
cntxgene <- rowSums(assays(spe)$counts)+1
plot.ecdf(cntxgene, xaxt="n", panel.first=grid(), xlab="UMI counts per gene",
          las=1, log="x", xlim=c(1, 10e5), main="")
axis(1, at=10^(0:5), labels=10^(0:5))
abline(v=100, lwd=2, col="red")
```
We filter out genes with less than 100 counts throughout all spots.

```{r}
dim(spe)
spe.filt <- spe[cntxgene > 100, ]
dim(spe.filt)
```

# Normalization

Calculate normalization factors and normalized expression values in logarithmic
scale.

```{r}
spe.filt <- computeLibraryFactors(spe.filt)
summary(sizeFactors(spe.filt))
spe.filt <- logNormCounts(spe.filt)
assayNames(spe.filt)
assays(spe.filt)$logcounts[1:5, 1:4]
```

Finally, store the quality controlled, normalized and processed data.

```{r}
saveRDS(spe.filt, file=file.path("_processed_data",
                                 "spe.filt.human_dlpfc.rds"))
```

# Gene Sets preparation

```{r, message=FALSE}
library(openxlsx)
library(dplyr)

cellmarkers <- read.xlsx(file.path("_annotation_data", "Cell_marker_Human.xlsx"),
                         rowNames=FALSE)
dim(cellmarkers)
head(cellmarkers, 3)
mask <- !is.na(cellmarkers$Symbol) & !is.na(cellmarkers$GeneID)
sum(mask)
cellmarkers <- cellmarkers[mask, ]
dim(cellmarkers)
mask <- cellmarkers$tissue_class == "Brain" &
        cellmarkers$cell_type == "Normal cell"
mask[is.na(mask)] <- FALSE
sum(mask)
braincellmarkers <- cellmarkers[mask, ]
head(sort(table(braincellmarkers$cell_name), decreasing=TRUE))
cell_types <- split(braincellmarkers$Symbol, braincellmarkers$cell_name)
length(cell_types)
```
Map gene identifiers to those in the `SpatialExperiment` object.

```{r}
cell_types <- lapply(cell_types,
                     function(x, speids, spesyms)
                       speids[spesyms %in% x],
                     rownames(spe.filt), rowData(spe.filt)$gene_name)
cell_types <- cell_types[lengths(cell_types) > 1]
length(cell_types)
names(cell_types) <- make.names(names(cell_types))
head(cell_types)
```

It is included in spatialLIBD (Pardo, Spangler, Weber et al., 2022) a set of genes from the Simons Foundation Autism Research Initiative SFARI.Below, we read in the data and create a list() object with the Ensembl gene IDs that SFARI has provided that are related to autism.


```{r}
## Read in the SFARI gene sets included in the package
asd_sfari <- utils::read.csv(
    system.file(
        "extdata",
        "SFARI-Gene_genes_01-03-2020release_02-04-2020export.csv",
        package = "spatialLIBD"
    ),
    as.is = TRUE
)

## Format them appropriately
asd_sfari_geneList <- list(
    Gene_SFARI_all = asd_sfari$ensembl.id,
    Gene_SFARI_high = asd_sfari$ensembl.id[asd_sfari$gene.score < 3],
    Gene_SFARI_syndromic = asd_sfari$ensembl.id[asd_sfari$syndromic == 1]
)
asd_sfari_geneList
```

Gene sets from Birnbaum et al. : 10 gene sets across SCZ, ASD, neurodevelopmental disorders, intellectual disability, bipolar disorder, and neurodegenerative disorders.

```{r}
sheet_data <- read.xlsx(file.path("_annotation_data","NIHMS613902-supplement-Supplement_Data.xlsx"),sheet = "Supp Table 1 Gene Sets")
df <- sheet_data[, 1:2]


cell_types_birnbaum <- split(df[,2],df[,1])
cell_types_birnbaum <- lapply(cell_types_birnbaum,
                     function(x, speids, spesyms)
                       speids[spesyms %in% x],
                     rownames(spe.filt), rowData(spe.filt)$gene_name)
cell_types_birnbaum <- cell_types_birnbaum[lengths(cell_types_birnbaum) > 1]
length(cell_types_birnbaum)
names(cell_types_birnbaum) <- make.names(names(cell_types_birnbaum))
head(cell_types_birnbaum)
```

We combine all gene lists and store the object.

```{r}
cell_types <- c(cell_types, asd_sfari_geneList,cell_types_birnbaum)
saveRDS(cell_types, file=file.path("_processed_data",
                                 "cell_types.human_dlpfc.rds"))

```
# Session information

```{r, child="_session-info.Rmd"}
```

# References
