---
title: GSVA analysis human DLPFC data
author: 
- name: Pere Moles
  affiliation:
  - &id Dept. of Medicine and Life Sciences, Universitat Pompeu Fabra
  email: pere.moles@upf.edu
date: "`r format(Sys.time(), '%B %e, %Y')`"
abstract: >
  Here we perform a GSVA analysis of the 10x Visium human DLPFC data.
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    fig_captions: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: _bibliography.bib
---

```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(kableExtra)


knitr::opts_chunk$set(
  collapse=TRUE,
  comment="",
  fig.align="center",
  fig.wide=TRUE,
  cache=TRUE,
  cache.path="_cache/__GS_hdlpfc",
  cache.extra=R.version.string,
  autodep=TRUE
)
```

# Import processed data

Start by importing processed data and gene sets.

```{r, message=FALSE}
library(SpatialExperiment)

spe.filt <- readRDS(file.path("_processed_data", "spe.filt.human_dlpfc.rds"))
dim(spe.filt)

cell_types <- readRDS(file.path("_processed_data", "cell_types.human_dlpfc.rds"))
```

# Dimensionality reduction

Select a subset of top highly-variable genes (HVGs).

```{r}
library(scran)

dec <- modelGeneVar(spe.filt, assay.type="logcounts")
```
Figure \@ref(fig:hdlpfcmeanvar) shows the mean-variance relationship.

```{r hdlpfcmeanvar, fig.height=4, fig.width=6, out.width="500px", fig.cap="Mean-variance relationship."}
fit <- metadata(dec)
plot(fit$mean, fit$var, xlab="mean of log-expression",
     ylab="variance of log-expression")
curve(fit$trend(x), col="dodgerblue", add=TRUE, lwd=2)
```
Select top 10% HVGs.

```{r}
top_hvgs <- getTopHVGs(dec, prop=0.1)
length(top_hvgs)
```
Calculate PCA.

```{r}
library(scater)

set.seed(123)
spe.filt <- runPCA(spe.filt, subset_row=top_hvgs)

reducedDimNames(spe.filt)
dim(reducedDim(spe.filt, "PCA"))
```

# Clustering

Here we perform a non-spatial graph-based clustering of the spots,
targeting 7 clusters, which is the number of anatomically annotated
groups of spots (six cortical layers and white matter) stored in the
column `spatialLIBD`.

```{r}
table(colData(spe.filt)$spatialLIBD)
k <- 19
g <- buildSNNGraph(spe.filt, k=k, use.dimred="PCA")
set.seed(123)
g_walk <- igraph::cluster_walktrap(g)
colLabels(spe.filt) <- factor(g_walk$membership)
table(colLabels(spe.filt))
```

Figure \@ref(fig:hdlpfcclustering) shows the clustering of the spots.

```{r hdlpfcclustering, message=FALSE, fig.height=5, fig.width=8, out.width="800px", dpi=100, fig.cap="Clustering of the spots."}
library(RColorBrewer)
library(patchwork)
library(ggspavis)

set.seed(123)
## image seems to be reversed and for this reason we set x_coord="pxl_col_in_fullres"
## and y_coord="pxl_row_in_fullres"
pltgt <- plotVisium(spe.filt, x_coord="pxl_col_in_fullres",
                    y_coord="pxl_row_in_fullres", point_size=0.75,
                    annotate="spatialLIBD",
                    pal=brewer.pal(nlevels(colLabels(spe.filt)), "Set1"))
pltcl <- plotVisium(spe.filt, x_coord="pxl_col_in_fullres",
                    y_coord="pxl_row_in_fullres", point_size=0.75, annotate="label",
                    pal=brewer.pal(nlevels(colLabels(spe.filt)), "Set1"))
(pltgt | pltcl)
```

# GSVA enrichment score estimation

Run GSVA with default parameters.

```{r, message=FALSE}
library(GSVA)
library(BiocParallel)

gsvaPar <- gsvaParam(spe.filt, cell_types, assay="logcounts",
                     minSize=5)
spe.filt.es <- gsva(gsvaPar, verbose=FALSE, BPPARAM=MulticoreParam(workers=3))
dim(spe.filt.es)
spe.filt.es
```

# Moran's I estimation

Here we calculate Moran's I for each gene set in the GSVA
enrichment scores (ES) calculated with default parameters.

```{r, message=FALSE}
library(SpatialFeatureExperiment)
library(Voyager)

sfe <- toSpatialFeatureExperiment(spe.filt.es)
vg <- findVisiumGraph(spe.filt.es,
                      sample_id="151673",
                      zero.policy=TRUE)
colGraph(sfe, "visium", sample_id="151673") <- vg

moransi <- rowData(runUnivariate(sfe, type="moran.test", colGraphName="visium",
                                 exprs_values="es", zero.policy=TRUE))
moransi <- moransi[, 2:3]
colnames(moransi) <- c("statistic", "p.value")
```


# Selection of autocorrelated gene sets

Table \@ref(tab:gsbymoransi) below shows the subset of gene sets with
significant nonzero Moran's I, in decreasing order, calculated from GSVA
ES with default parameters.

```{r}
mask <- p.adjust(moransi$p.value, method="bonferroni") < 0.01
sum(mask)
moransi <- moransi[mask, ]
dim(moransi)
```

```{r gsbymoransi, echo=FALSE}
ktab <- kable(moransi[order(moransi$statistic, decreasing=TRUE), ], "html",
              escape=FALSE, row.names=TRUE,
              caption="Significantly autocorrelated gene sets, ordered by decreasing Morans'I.")
ktab <- kable_styling(ktab,
                      bootstrap_options=c("striped", "hover", "responsive"),
                      fixed_thead=TRUE)
kable_styling(ktab, position="center")
```
We subset the GSVA ES calculated with default parameters to those gene sets
significantly autocorrelated with a Bonferroni FWER < 0.01, and a minimum
Moran's I value of 0.4.

```{r}
sig.genesets <- rownames(moransi)[moransi$statistic > 40]
spe.filt.es.sig <- spe.filt.es[rownames(spe.filt.es) %in% sig.genesets]
dim(spe.filt.es.sig)
```

# Cluster annotation

Here we use the previously selected autocorrelated gene sets to annotate
the gene-level clusters. We first define the following function for that
purpose, which takes as input an `SpatialExperiment` object with clusters
annotated in the `label` column metadata, and return a vector with as many
values as clusters and the most similar gene set for each cluster, in the
corresponding position.

```{r}
clustertogeneset <- function(spe, p.value=0.01) {
  require(limma)

  y <- assay(spe)
  f <- colData(spe)$spatialLIBD
  stopifnot(is.factor(f) && length(f) > 1) ## QC
  nl <- nlevels(f)
  mod <- model.matrix(~ 0 + f)
  genesets <- c()
  for (targetcluster in levels(f)) {
    fit <- lmFit(y, mod)
    ## compare mean of the target cluster against the mean of
    ## all other clusters
    cf <- sprintf("f%s=f%s-(%s)/%s", targetcluster, targetcluster,
                  paste(sprintf("f%s", setdiff(levels(f), targetcluster)),
                        collapse="+"),
                  nl-1)
    contrast.matrix <- makeContrasts(contrasts=cf, levels=mod)
    fit <- contrasts.fit(fit, contrast.matrix)
    fit <- eBayes(fit)
    tt <- topTable(fit, coef=1, n=Inf)
    tt <- tt[tt$adj.P.Val < p.value, , drop=FALSE]
    genesets <- c(genesets, rownames(tt)[1])
  }
  genesets
}
```
Annotate gene sets to clusters and keep the unique ones using the GSVA ES
calculated with default parameters.

```{r, message=FALSE}
annotatedgs <- clustertogeneset(spe.filt.es.sig)
annotatedgs
annotatedgs <- unique(annotatedgs)
annotatedgs
```
Figure \@ref(fig:hcannotclusters) shows these unique gene sets that mostly
correlated with some of the clusters.

```{r hcannotclusters, echo=FALSE, message=FALSE, fig.height=10, fig.width=7, out.width="800px", dpi=100, fig.cap="Annotation of clusters with gene sets autocorrelated with GSVA ES calculated with default parameters. (a) Original tissue image. (b) Spot annotation by gene-level clustering. (c) to (h) enrichment scores for each of the significantly autocorrelated gene sets that mostly correlated with at least one of the gene-level clusters."}
library(ggspavis)
library(patchwork)
library(RColorBrewer)

set.seed(123)
plts <- list(
  original = plotVisium(spe.filt,
                        x_coord="pxl_col_in_fullres",
                        y_coord="pxl_row_in_fullres",
                        point_size=0.25, facets=NULL) +
            ggtitle("Original tissue") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            labs(tag = "a"),
  annotated = plotVisium(spe.filt, 
                        x_coord="pxl_col_in_fullres",
                        y_coord="pxl_row_in_fullres", 
                        point_size = 1,
                        annotate = "spatialLIBD", 
                        facets = NULL,
                        pal = brewer.pal(nlevels(colData(spe.filt)$spatialLIBD), "Set1")
                        ) +
            ggtitle("Ground truth labels") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            labs(tag = "b")
)

i <- 3
for (gs in annotatedgs) {
  plts[[gs]] <- plotVisium(spe.filt.es.sig, x_coord="pxl_col_in_fullres",
                           y_coord="pxl_row_in_fullres", point_size=0.75,
                           assay="es", annotate=gs, facets=NULL) +
                      ggtitle(gs) + theme(plot.title=element_text(hjust=0.5)) +
                      theme(legend.title=element_blank()) + labs(tag=letters[i])
  i <- i + 1
}
wrap_plots(plts, nrow=ceiling(length(plts)/2))
```

Finally, store the GSVA ES calculated with default parameters.
```{r}
saveRDS(spe.filt.es,
        file=file.path("_processed_data",
                       "spe.filt.es.human_dlpfc.rds"))
```


## Plot SFARI gene sets
Figure \@ref(fig:sfarigenesets) shows GSVA scores for SFARI gene sets.

```{r sfarigenesets, echo=FALSE, message=FALSE, fig.height=10, fig.width=7, out.width="800px", dpi=100, fig.cap="GSVA scores for SFARI gene sets"}


set.seed(123)

plts_sfari <- list()
for (gs in names(cell_types[c(96:98)])) {
  plts_sfari[[gs]] <- plotVisium(spe.filt.es, x_coord="pxl_col_in_fullres",
                           y_coord="pxl_row_in_fullres", point_size=0.75,
                           assay="es", annotate=gs, facets=NULL) +
                      ggtitle(gs) + theme(plot.title=element_text(hjust=0.5)) +
                      theme(legend.title=element_blank()) + labs(tag=letters[i])
  i <- i + 1
}
wrap_plots(plts_sfari, nrow=ceiling(length(plts)/2))
```
## Plot Birnbaum gene sets

Figure \@ref(fig:birnbaumgenesets) shows GSVA scores for Birnbaum et al gene sets.

```{r birnbaumgenesets, echo=FALSE, message=FALSE, fig.height=10, fig.width=7, out.width="800px", dpi=100, fig.cap="GSVA scores for Birnbaum et al gene sets."}


set.seed(123)

plts_birnbaum <- list()
for (gs in names(cell_types[c(99:108)])) {
  plts_birnbaum[[gs]] <- plotVisium(spe.filt.es, x_coord="pxl_col_in_fullres",
                           y_coord="pxl_row_in_fullres", point_size=0.75,
                           assay="es", annotate=gs, facets=NULL) +
                      ggtitle(gs) + theme(plot.title=element_text(hjust=0.5)) +
                      theme(legend.title=element_blank()) + labs(tag=letters[i])
  i <- i + 1
}
wrap_plots(plts_birnbaum, nrow=ceiling(length(plts)/2))
```

# Session information

```{r, child="_session-info.Rmd"}
```

# References
