---
title: GSVA analysis mouse brain data
author: 
- name: Pere Moles
  affiliation:
  - &id MSc Data Science student, Universitat Oberta de Catalunya
  email: pere.moles@uoc.edu

date: "`r format(Sys.time(), '%B %e, %Y')`"
abstract: >
  Here we perform a GSVA analysis of the mouse brain data.
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    fig_captions: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: _bibliography.bib
---

```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(kableExtra)


knitr::opts_chunk$set(
  collapse=TRUE,
  comment="",
  fig.align="center",
  fig.wide=TRUE,
  cache=TRUE,
  cache.path="_cache/__GS_mouse",
  cache.extra=R.version.string,
  autodep=TRUE
)
```

# Import processed data

Start by importing processed data.

```{r, message=FALSE}
library(SpatialExperiment)

spe.filt <- readRDS(file.path("_processed_data", "spe.filt.mouse_brain.rds"))
dim(spe.filt)

cell_types <- readRDS(file.path("_processed_data", "cell_types.mouse_brain.rds"))
```

# Dimensionality reduction

Select a subset of top highly-variable genes (HVGs).

```{r}
library(scran)

dec <- modelGeneVar(spe.filt, assay.type="logcounts")
```
Figure \@ref(fig:mousemeanvar) shows the mean-variance relationship.

```{r mousemeanvar, fig.height=4, fig.width=6, out.width="500px", fig.cap="Mean-variance relationship."}
fit <- metadata(dec)
plot(fit$mean, fit$var, xlab="mean of log-expression",
     ylab="variance of log-expression")
curve(fit$trend(x), col="dodgerblue", add=TRUE, lwd=2)
```
Select top 10% HVGs.

```{r}
top_hvgs <- getTopHVGs(dec, prop=0.1)
length(top_hvgs)
```
Calculate PCA.

```{r}
library(scater)

set.seed(123)
spe.filt <- runPCA(spe.filt, subset_row=top_hvgs)

reducedDimNames(spe.filt)
dim(reducedDim(spe.filt, "PCA"))
```

# Clustering

Here we perform a non-spatial graph-based clustering of the spots.

```{r}
k <- 30
g <- buildSNNGraph(spe.filt, k=k, use.dimred="PCA")
set.seed(123)
g_walk <- igraph::cluster_walktrap(g)
colLabels(spe.filt) <- factor(g_walk$membership)
table(colLabels(spe.filt))
```

Figure \@ref(fig:mouseclustering) shows the clustering of the spots.

```{r mouseclustering, message=FALSE, fig.height=5, fig.width=8, out.width="800px", dpi=100, fig.cap="Clustering of the spots."}
library(RColorBrewer)
library(patchwork)
library(ggspavis)

set.seed(123)
pltcl <- plotVisium(spe.filt, x_coord="pxl_row_in_fullres",
                    y_coord="pxl_col_in_fullres", point_size=1, annotate="label",
                    pal=brewer.pal(nlevels(colLabels(spe.filt)), "Set1"))
pltcl
```


# GSVA enrichment score estimation

Run GSVA with default parameters.

```{r, message=FALSE}
library(GSVA)
library(BiocParallel)

gsvaPar <- gsvaParam(spe.filt, cell_types, assay="logcounts", kcdf = 'none',
                     minSize=5)
spe.filt.es <- gsva(gsvaPar, verbose=FALSE, BPPARAM=MulticoreParam(workers=3))
dim(spe.filt.es)
spe.filt.es
```


# Moran's I estimation

Here we calculate Moran's I for each gene set in the GSVA
enrichment scores (ES) calculated with default parameters.

```{r, message=FALSE}
library(SpatialFeatureExperiment)
library(Voyager)

sfe <- toSpatialFeatureExperiment(spe.filt.es)
vg <- findVisiumGraph(spe.filt.es,
                      sample_id="MouseBrainSagittalAnterior1",
                      zero.policy=TRUE)
colGraph(sfe, "visium", sample_id="MouseBrainSagittalAnterior1") <- vg

moransi <- rowData(runUnivariate(sfe, type="moran.test", colGraphName="visium",
                                 exprs_values="es", zero.policy=TRUE))
moransi <- moransi[, 2:3]
colnames(moransi) <- c("statistic", "p.value")
```


# Selection of autocorrelated gene sets

Table \@ref(tab:gsbymoransi) below shows the subset of gene sets with
significant nonzero Moran's I, in decreasing order, calculated from GSVA
ES with default parameters.

```{r}
mask <- p.adjust(moransi$p.value, method="bonferroni") < 0.01
sum(mask)
moransi <- moransi[mask, ]
dim(moransi)
```

```{r gsbymoransi, echo=FALSE}
ktab <- kable(moransi[order(moransi$statistic, decreasing=TRUE), ], "html",
              escape=FALSE, row.names=TRUE,
              caption="Significantly autocorrelated gene sets, ordered by decreasing Morans'I.")
ktab <- kable_styling(ktab,
                      bootstrap_options=c("striped", "hover", "responsive"),
                      fixed_thead=TRUE)
kable_styling(ktab, position="center")
```
We subset the GSVA ES calculated with default parameters to those gene sets
significantly autocorrelated with a Bonferroni FWER < 0.01, and a minimum
Moran's I value of 0.4.

```{r}
sig.genesets <- rownames(moransi)[moransi$statistic > 40]
spe.filt.es.sig <- spe.filt.es[rownames(spe.filt.es) %in% sig.genesets]
dim(spe.filt.es.sig)
```

# Cluster annotation

Here we use the previously selected autocorrelated gene sets to annotate
the gene-level clusters. We first define the following function for that
purpose, which takes as input an `SpatialExperiment` object with clusters
annotated in the `label` column metadata, and return a vector with as many
values as clusters and the most similar gene set for each cluster, in the
corresponding position.

```{r}
clustertogeneset <- function(spe, p.value=0.01) {
  require(limma)

  y <- assay(spe)
  f <- colData(spe)$label
  stopifnot(is.factor(f) && length(f) > 1) ## QC
  nl <- nlevels(f)
  mod <- model.matrix(~ 0 + f)
  genesets <- c()
  for (targetcluster in levels(f)) {
    fit <- lmFit(y, mod)
    ## compare mean of the target cluster against the mean of
    ## all other clusters
    cf <- sprintf("f%s=f%s-(%s)/%s", targetcluster, targetcluster,
                  paste(sprintf("f%s", setdiff(levels(f), targetcluster)),
                        collapse="+"),
                  nl-1)
    contrast.matrix <- makeContrasts(contrasts=cf, levels=mod)
    fit <- contrasts.fit(fit, contrast.matrix)
    fit <- eBayes(fit)
    tt <- topTable(fit, coef=1, n=Inf)
    tt <- tt[tt$adj.P.Val < p.value, , drop=FALSE]
    genesets <- c(genesets, rownames(tt)[1])
  }
  genesets
}
```
Annotate gene sets to clusters and keep the unique ones using the GSVA ES
calculated with default parameters.

```{r, message=FALSE}
annotatedgs <- clustertogeneset(spe.filt.es.sig)
annotatedgs
annotatedgs <- unique(annotatedgs)
annotatedgs
```
Figure \@ref(fig:hcannotclusters) shows these unique gene sets that mostly
correlated with some of the clusters.

```{r hcannotclusters, echo=FALSE, message=FALSE, fig.height=10, fig.width=7, out.width="800px", dpi=100, fig.cap="Annotation of clusters with gene sets autocorrelated with GSVA ES calculated with default parameters. (a) Original tissue image. (b) Spot annotation by gene-level clustering. (c) to (h) enrichment scores for each of the significantly autocorrelated gene sets that mostly correlated with at least one of the gene-level clusters."}
library(ggspavis)
library(patchwork)
library(RColorBrewer)

set.seed(123)
plts <- list(
  original = plotVisium(spe.filt,
                        x_coord="pxl_row_in_fullres",
                        y_coord="pxl_col_in_fullres",
                        point_size=0.25, facets=NULL) +
            ggtitle("Original tissue") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            labs(tag = "a"),
  annotated = plotVisium(spe.filt, 
                        x_coord="pxl_row_in_fullres",
                        y_coord="pxl_col_in_fullres", 
                        point_size = 1,
                        annotate = "label", 
                        facets = NULL,
                        pal = brewer.pal(nlevels(colData(spe.filt)$label), "Set1")
                        ) +
            ggtitle("Cluster labels") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            labs(tag = "b")
)

i <- 3
for (gs in annotatedgs) {
  plts[[gs]] <- plotVisium(spe.filt.es.sig, x_coord="pxl_row_in_fullres",
                           y_coord="pxl_col_in_fullres", point_size=0.75,
                           assay="es", annotate=gs, facets=NULL) +
                      ggtitle(gs) + theme(plot.title=element_text(hjust=0.5)) +
                      theme(legend.title=element_blank()) + labs(tag=letters[i])
  i <- i + 1
}
wrap_plots(plts, nrow=ceiling(length(plts)/2))
```
Annotate gene sets to clusters and keep the unique ones using the GSVA ES
calculated without kernel estimation.

Finally, store the GSVA ES calculated with default parameters and without kernel
estimation.

```{r}
saveRDS(spe.filt.es,
        file=file.path("_processed_data",
                       "spe.filt.es.mouse.rds"))
```


# Session information

```{r, child="_session-info.Rmd"}
```

# References








